---
title: "The Multiverse"
subtitle: "Every line is a justifiable analysis"
engine: knitr
format:
  live-html:
    toc: true
    webr:
      packages:
        - ggplot2
resources:
  - data/spec_grid.csv
  - data/analysts5.csv
  - data/class_results.csv
execute:
  echo: false
  warning: false
  message: false
---

{{< include _extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| label: setup
#| include: false
# Palette mirrors slides/index.qmd setup so this page and the deck share one
# visual grammar: red = the reproduced baseline, orange = you / C6HJR, grey = the
# other analysts. Static charts below are rendered by knitr at build time; the
# live class-results chart further down runs in the browser via webR.
library(ggplot2)
ncl_blue <- "#0e3b6a"; orange <- "#E69F00"; green <- "#009E73"
d_red    <- "#a10000"; grey_mid <- "#848c94"
```

This page shows the specification curve for the workshop exercise. The static charts below render at build time from the pre-computed menu; the **Class results** chart at the foot of the page is live – it fetches the room's submissions in your browser and redraws when you press *Run Code*. Before the session – or when the data files are absent – the page renders with placeholder messages.

The logic of the visualisation follows the specification-curve idea (Simonsohn, Simmons and Nelson): one point per justifiable analysis, ordered by size. Because the menu mixes outcome scales and exposure transforms, raw coefficients are not comparable across specifications – the chart therefore plots the **partial correlation** (r = t/sqrt(t^2 + df)), the same metric Multi100 used to compare analysts. A horizontal line at zero marks the boundary between negative and positive associations. Your contribution sits somewhere on this chart.

---

## Specification curve

```{r}
#| label: spec-curve
#| fig-width: 9
#| fig-height: 5
#| fig-alt: "Specification curve: partial correlations for 1,680 specifications, ordered by size, coloured by outcome variable, with the constrained baseline marked. Horizontal line at zero."

library(ggplot2)

spec_file <- "data/spec_grid.csv"

if (file.exists(spec_file)) {
  spec <- read.csv(spec_file)

  required_cols <- c("estimate", "statistic", "df", "r", "outcome", "baseline")
  if (all(required_cols %in% names(spec))) {

    # order by partial r - the cross-specification comparable metric
    spec <- spec[order(spec$r), ]
    spec$rank <- seq_len(nrow(spec))
    spec$sig <- abs(spec$statistic) > qt(0.975, spec$df)
    base_pt <- spec[spec$baseline, ]

    p <- ggplot(spec, aes(x = rank, y = r)) +
      geom_hline(yintercept = 0, colour = grey_mid, linewidth = 0.4) +
      geom_point(aes(colour = outcome, alpha = sig), size = 1.2) +
      geom_point(data = base_pt, shape = 23, size = 3.5,
                 fill = d_red, colour = d_red, stroke = 0) +
      annotate("text", x = base_pt$rank, y = base_pt$r - 0.08,
               label = "constrained baseline\n(t = -3.80)", size = 3,
               colour = d_red, hjust = 0.2) +
      scale_colour_brewer(palette = "Dark2", name = "Outcome") +
      scale_alpha_manual(values = c("FALSE" = 0.25, "TRUE" = 0.9),
                         name = "p < .05", labels = c("no", "yes")) +
      labs(
        x = "Specification (ordered by partial r)",
        y = "Partial correlation (unemployment coefficient)",
        title = "1,680 justifiable analyses of one claim",
        caption = "Under the claim, positive framings (mcosmo, mutil, mpos) should sit below zero and negative framings (mcomm, mlib, mneg) above it. Faded points: not significant at 5%."
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", colour = ncl_blue),
        axis.title = element_text(colour = ncl_blue),
        legend.position = "bottom"
      )

    print(p)

  } else {
    cat("**Note:** `data/spec_grid.csv` is present but missing expected columns (",
        paste(setdiff(required_cols, names(spec)), collapse = ", "),
        "). Chart will render once the file is complete.\n")
  }

} else {
  cat("**Note:** `data/spec_grid.csv` not yet available. The specification curve will appear here once the pre-computed multiverse file is in place.\n")
}
```

```{r}
#| label: spec-summary
#| output: asis

if (file.exists(spec_file)) {
  s <- read.csv(spec_file)
  s$sig <- abs(s$statistic) > qt(0.975, s$df)
  cat(sprintf(
    paste0(
      "Across all %d specifications: **%.0f%%** point in the claim-consistent direction, ",
      "**%.0f%%** are significant at the 5%% level, and **%.0f%%** are both. ",
      "The partial correlation ranges from %.2f to %.2f. ",
      "Same data, same claim – the answer depends on the path.\n"
    ),
    nrow(s), 100 * mean(s$supports_claim), 100 * mean(s$sig),
    100 * mean(s$sig & s$supports_claim), min(s$r), max(s$r)
  ))
}
```

---

## Five Multi100 analysts (partial r)

```{r}
#| label: analysts-plot
#| fig-width: 7
#| fig-height: 3.2
#| fig-alt: "Dot plot of partial r values for the five Multi100 analysts. Horizontal line at zero."

analysts_file <- "data/analysts5.csv"

if (file.exists(analysts_file)) {
  analysts <- read.csv(analysts_file)

  if ("r" %in% names(analysts) && "analyst_id" %in% names(analysts)) {

    # Highlight the workshop analyst
    analysts$highlight <- analysts$analyst_id == "C6HJR"

    p2 <- ggplot(analysts, aes(x = r, y = reorder(analyst_id, r))) +
      geom_vline(xintercept = 0, colour = grey_mid, linewidth = 0.4) +
      geom_point(
        aes(colour = highlight, size = highlight)
      ) +
      scale_colour_manual(values = c("FALSE" = grey_mid, "TRUE" = orange),
                          guide = "none") +
      scale_size_manual(values = c("FALSE" = 2.5, "TRUE" = 3.5), guide = "none") +
      labs(
        x = "Partial r",
        y = NULL,
        title = "Five Multi100 analysts – Task 2 partial r",
        caption = "Orange point = analyst C6HJR"
      ) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", colour = ncl_blue),
            axis.title = element_text(colour = ncl_blue))

    print(p2)

  } else {
    cat("**Note:** `data/analysts5.csv` is present but missing `r` or `analyst_id` column. Chart will render once the file is complete.\n")
  }

} else {
  cat("**Note:** `data/analysts5.csv` not yet available. The analyst plot will appear here once the file is in place.\n")
}
```

---

## Class results – 16 June 2026

This chart is **live**. It draws the pre-computed menu (grey) and the five Multi100 analysts (the dashed lines, C6HJR in orange) as a backdrop, then overlays the room's own submissions as orange dots, placed where each result's partial *r* falls on the curve. Press **Run Code** to pull the latest submissions.

The R engine downloads into your browser the first time (10–30 seconds); the button changes from a spinner to *Run Code* when it is ready.

```{webr}
#| label: live-multiverse
#| autorun: true
#| fig-width: 9
#| fig-height: 5
#| caption: "Press Run Code to refresh – pull the latest class results"
# ---------------------------------------------------------------------------
# CONFIG. Set worker_url to your Cloudflare Worker URL once it is deployed
# (see facilitator/live-results-setup.md, steps 1-3). Leave it as "" before the
# Worker exists: the chart still renders the backdrop + the five analysts, so the
# page works at any stage of setup.
worker_url <- "https://class-results.chris-moreh.workers.dev/"
# ---------------------------------------------------------------------------

library(ggplot2)
ncl_blue <- "#0e3b6a"; orange <- "#E69F00"; d_red <- "#a10000"; grey_mid <- "#848c94"

# Backdrop, from the page's bundled resources (always available, even with zero
# submissions or a failed fetch).
spec <- read.csv("data/spec_grid.csv")
spec <- spec[order(spec$r), ]
spec$rank <- seq_len(nrow(spec))
an <- read.csv("data/analysts5.csv")
an$me <- an$analyst_id == "C6HJR"

# Class results. Three regimes, in order of preference:
#   1. live  – fetch from the Worker (cache-busted so each Run pulls fresh);
#   2. archive – if the Worker is unset or fails, read a committed snapshot
#                (data/class_results.csv), written after the session;
#   3. none  – fall back to just the backdrop.
read_clean <- function(path) {
  d <- read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
  names(d) <- tolower(trimws(names(d)))   # tolerate "R", " r ", etc.
  d$r <- suppressWarnings(as.numeric(d$r))
  d[!is.na(d$r), , drop = FALSE]
}
cls <- NULL
if (nzchar(worker_url)) {
  cls <- tryCatch({
    download.file(paste0(worker_url, "?_=", as.numeric(Sys.time())),
                  "class.csv", quiet = TRUE)
    read_clean("class.csv")
  }, error = function(e) NULL)
}
if (is.null(cls) && file.exists("data/class_results.csv")) {
  cls <- tryCatch(read_clean("data/class_results.csv"), error = function(e) NULL)
}

p <- ggplot(spec, aes(rank, r)) +
  geom_hline(yintercept = 0, colour = grey_mid, linewidth = 0.4) +
  geom_point(colour = "#c7ccd2", size = 0.7, alpha = 0.5) +
  geom_hline(data = an[!an$me, ], aes(yintercept = r),
             colour = grey_mid, linetype = "22", linewidth = 0.6) +
  geom_hline(data = an[an$me, ], aes(yintercept = r),
             colour = orange, linetype = "22", linewidth = 0.9) +
  labs(x = "specification (ranked by partial r)", y = "partial r",
       title = "The class multiverse – live") +
  coord_cartesian(ylim = c(-0.75, 0.78)) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", colour = ncl_blue),
        axis.title = element_text(colour = ncl_blue))

if (!is.null(cls) && nrow(cls) > 0) {
  cls$rank <- findInterval(cls$r, spec$r)
  p <- p +
    geom_point(data = cls, aes(rank, r), colour = orange, size = 3, alpha = 0.9) +
    labs(subtitle = paste(nrow(cls), "class result(s) so far"))
} else {
  p <- p + labs(subtitle = "No submissions yet – press Run Code to refresh")
}
p
```

<!-- ::: {.callout-note collapse="true"}
## How this works (facilitator)

Submissions go to a Google Form, land in a linked Google Sheet, and are served as CSV with permissive CORS headers by a small Cloudflare Worker. The `{webr}` cell above fetches that CSV in the browser. 
::: -->
