---
title: "Missing answers"
subtitle: "The respondents who mentioned nothing, and the conventions for them"
bibliography: [../references.bib, ../references-missing.bib]
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({ library(dplyr); library(ggplot2); library(fixest) })
cy  <- datawizard::data_read("../data/EUframes_cy.csv", verbose = FALSE)
pw  <- datawizard::data_read("../data/teney_panel_weights.csv", verbose = FALSE)
me  <- datawizard::data_read("../data/teney_panel_measurement.csv", verbose = FALSE)
mi  <- datawizard::data_read("../data/spec_grid_mi.csv", verbose = FALSE)
dg  <- datawizard::data_read("../data/mi_diagnostics.csv", verbose = FALSE)
lev <- datawizard::data_read("../data/axis_level_summary_maximal.csv", verbose = FALSE)
cell <- readr::read_csv("../data/grid/mcosmo__unemp__none__fe_twoway__all__none.csv.gz",
                        show_col_types = FALSE)
pct <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
num <- function(x, d = 3) gsub("-", "−", formatC(x, digits = d, format = "f"), fixed = TRUE)
big <- function(x) format(x, big.mark = ",")

n_all  <- sum(pw$n_cy); n_noop <- sum(pw$n_noop)
n_dk <- sum(pw$n_dk); n_oth <- sum(pw$n_other); n_bl <- sum(pw$n_blank)
r_noop_unemp <- cor(pw$noop_rate, cy$unemp[match(paste(pw$cntry, pw$year), paste(cy$cntry, cy$year))])
r_noop_cosmo <- cor(pw$noop_rate, cy$mcosmo[match(paste(pw$cntry, pw$year), paste(cy$cntry, cy$year))])
by_c <- pw |> summarise(rate = mean(noop_rate), .by = cntry) |> arrange(desc(rate))

# the published specification under each treatment of the no-opinion respondents
spec <- cell |>
  filter(construction == "share", item_set == "all", cell_weight == "unweighted",
         measure == "wb_nat", pred_form == "raw", se == "conventional") |>
  select(noop, estimate, statistic, df, n, p, r_ca) |>
  mutate(noop = factor(noop, c("zero", "drop", "split", "mi_mice", "mi_amelia")))
mi_spec <- mi |> filter(outcome == "mcosmo", copredictor == "none", estimator == "fe_twoway",
                        sample == "all", reg_weight == "none", se == "conventional")
noop_lev <- lev |> filter(axis == "noop")

# the sensitivity band: every silent respondent is assigned the answered mean of
# their cell plus delta; delta = 0 is the drop rule and delta = -(cell mean) is
# the zero rule, so the two conventions are the two ends of one assumption
ans <- me |> filter(construction == "share", item_set == "all", noop == "drop",
                    cell_weight == "unweighted", outcome == "mcosmo") |>
  select(cntry, year, m_ans = value) |>
  inner_join(pw |> select(cntry, year, n_cy, n_noop), by = c("cntry", "year")) |>
  inner_join(cy |> select(cntry, year, unemp), by = c("cntry", "year")) |>
  mutate(share_noop = n_noop / n_cy)
band <- purrr::map(seq(-0.6, 0.3, by = 0.05), function(d) {
  m <- feols(I(m_ans + share_noop * d) ~ unemp | cntry + year, data = ans)
  tibble(delta = d, b = coef(m)[["unemp"]], t = coeftable(m)["unemp", "t value"])
}) |> purrr::list_rbind()
delta_zero <- -weighted.mean(ans$m_ans, ans$n_noop)
```

The [specification menu](../exercise/spec-menu.qmd) carries an axis called **no-opinion
handling** with four settings, and this module is the argument for why it is there. The
short version is that `r big(n_noop)` of the `r big(n_all)` respondents in the EU-frames
data, `r pct(n_noop / n_all)`, answered the question "What does the EU mean to you
personally?" by mentioning none of the thirteen things on the list, so that on every
framing scale their score is zero divided by zero. The original paper and its five
reanalyses resolved that in two different ways without saying so, and the people
concerned are not a random slice of the sample.

## The missing answers

The framing scales are built from a multiple-mention battery: a respondent ticks any of
thirteen items, and the cosmopolitan scale is the share of their ticks that fall on the
five cosmopolitan items. A respondent who ticks nothing has no share. The survey records
two further codes for that person, "other (spontaneous)" and "don't know", and the
`r big(n_noop)` split three ways: `r big(n_dk)` said they did not know, `r big(n_oth)`
gave an answer that was not on the list, and `r big(n_bl)` have no tick of any kind.
Those are three different people. Someone who says the EU means something the list
did not anticipate has an opinion; someone who says they do not know may have none, or
may have one they would rather not state; a blank row is a refusal, an interviewer's
omission, or a coding gap, and the file does not say which.

The original author's Stata code divides each cluster count by the total of mentions and
has no rule for a zero total. Stata returns a missing value for 0/0, and the mixed model
drops a missing outcome, so those respondents left her analysis without a line of code
saying so: her reported level-1 N of 390,373 is exactly the number of respondents who
mentioned at least one item. The Multi100 reanalysis that produced the recorded t of
−3.804 coded the same respondents 0 on every scale, and the course panel inherits that
convention. One analyst, PRL47, built both versions and reported the zero one. So the
published coefficient and its best-known reanalysis differ in what they assume about
`r pct(n_noop / n_all)` of the sample, and neither states the assumption.

Whether that matters depends on who the silent respondents are, and here the answer is
unambiguous. They are older, more often women, and they live in country-years with
higher unemployment. Across the `r nrow(pw)` cells the no-opinion rate correlates at
`r num(r_noop_unemp, 2)` with the unemployment rate and at `r num(r_noop_cosmo, 2)`
with the cosmopolitan scale. The rate also differs sharply between countries: it runs at
`r pct(by_c$rate[1])` in `r by_c$cntry[1]` and `r pct(by_c$rate[2])` in `r by_c$cntry[2]`
against `r pct(by_c$rate[nrow(by_c)])` in `r by_c$cntry[nrow(by_c)]`. Scoring the silent
as zero therefore pulls the framing scale down most where unemployment is highest, which
is the direction the claim predicts. Part of the association is a coding rule.

## The three missingness mechanisms

The vocabulary for thinking about this comes from @Rubin1976InferenceMissingData and is
set out accessibly by @Dong2013Principled. Data are **missing completely at random**
when the probability that a value is missing depends on nothing in the data at all, so
the observed cases are a random sample of the complete ones; dropping them costs
precision and nothing else. Data are **missing at random** when that probability depends
on things that were observed but not, once those are accounted for, on the missing value
itself: for students who scored alike on a pre-test, in the example Dong and Peng use,
whether the post-test is missing is a coin toss. Data are **missing not at random** when
the probability depends on the missing value even after the observed data are allowed
for, as when high earners are the ones who decline to state their income. The labels
describe the mechanism that produced the gaps, not their size, and the practical
consequence is asymmetric: the first two can be handled with methods that use only the
observed data, while the third requires the analyst to say how the missing values differ
from the observed ones, which the data cannot reveal.

Mapped onto the case, the silent respondents are plainly not missing completely at
random, since their age, sex and location predict their silence. Missing at random given
age, sex, education and the country-year cell is the working assumption that every
imputation below rests on. It is not testable from the observed data alone, a point
@Dong2013Principled make against the habit of reading a t-test on observed covariates
as a test of it; what can be done is to make it more plausible by conditioning on
everything that predicts silence, and then to ask how far the conclusion moves if it is
wrong. That last step is the sensitivity band at the end of this module. A "don't know"
about what the EU means is a fair candidate for missing not at random, because the
people least engaged with the EU are the people least likely to have a framing of it,
and disengagement is not in the data.

## Three levels

Item non-response on the outcome battery is the case this module works, but the same
structure recurs at two other levels of this analysis, and the [weighting
module](weighting.qmd) covers the other two. Inside a country, some people are harder to
reach or more likely to refuse the whole interview, which is unit non-response, and the
Eurobarometer W1 design weight is the standard correction for the imbalance it leaves.
Above the country, one reanalyst's Eurostat table had no United Kingdom row and nothing
before 2009 for any country but France, so half the country-year cells were missing from
that analysis before a single respondent was considered. Nonresponse rates on their own
predict little about the bias that results [@Groves2006Nonresponse; @Groves2008Impact;
@Hendra2018Rethinking]; what matters is whether the missing units differ from the
observed ones on the quantity of interest, which is the same question at every level.

## The conventions

Three deterministic rules are on the menu, and each is a statement about what the silent
respondents mean. **Zero** treats them as people with no framing of any kind, which is
the reanalysis convention. **Drop** treats them as people who did not answer the
question, which is what the original author's code did. **Split** takes the three codes
at their word: "don't know" and blank are dropped as non-response, and "other" is kept
at zero on the reasoning that an answer off the list is still an answer that contains no
cosmopolitan content. The rules change the cell means before any model is fitted: the
average cosmopolitan cell mean is `r num(mean(me$value[me$construction == "share" & me$item_set == "all" & me$noop == "zero" & me$cell_weight == "unweighted" & me$outcome == "mcosmo"]), 3)`
under zero and `r num(mean(me$value[me$construction == "share" & me$item_set == "all" & me$noop == "drop" & me$cell_weight == "unweighted" & me$outcome == "mcosmo"]), 3)`
under drop.

The rules also change the question. Under zero the estimand is the average framing among
everyone interviewed, with silence counted as no framing; under drop it is the average
framing among people who have a framing at all. Those are different populations, and by
the test the [specification menu](../exercise/spec-menu.qmd#what-each-decision-commits-you-to)
applies, that makes the no-opinion rule an estimand decision dressed as a measurement
one. It is classed with measurement on the menu because that is where the decision is
made in practice, at the moment the outcome column is built, and a participant meeting
it there should know that it reaches further.

Across the whole grid the rule moves the claim-direction share and the significant share
by a few points and the median correlation by a few hundredths:

```{r}
#| label: tbl-noop-levels
#| echo: false
tinytable::tt(
  noop_lev |>
    mutate(level = factor(level, c("zero", "drop", "split", "mi_mice", "mi_amelia"),
                          labels = c("zero", "drop", "split", "imputed (mice)", "imputed (Amelia)"))) |>
    arrange(level) |>
    transmute(`No-opinion rule` = level, Specifications = big(n),
              `Point the claim's way` = pct(claim_share / 100),
              Significant = pct(sig_share / 100),
              `Median aligned r` = num(median_r_ca)) |>
    as.data.frame()) |>
  tinytable::style_tt(j = 2:5, align = "r") |>
  tinytable::style_tt(fontsize = 0.8)
```

: The no-opinion rule across every specification in which it varies. The two imputation rows are fitted at the canonical measurement settings only. {#tbl-noop-levels}

The published specification shows what those averages conceal. Under the five rules the
coefficient on unemployment for cosmopolitan framing, with two-way fixed effects on all
270 cells, is:

```{r}
#| label: tbl-noop-spec
#| echo: false
tinytable::tt(
  spec |> arrange(noop) |>
    transmute(Rule = c(zero = "zero (Multi100)", drop = "drop (original paper)",
                       split = "split", mi_mice = "imputed, mice",
                       mi_amelia = "imputed, Amelia")[as.character(noop)],
              b = gsub("-", "−", formatC(estimate, digits = 3, format = "g"), fixed = TRUE),
              t = num(statistic, 2), df = formatC(df, format = "f", digits = 0),
              p = gsub("-", "−", formatC(p, digits = 2, format = "g"), fixed = TRUE),
              `aligned r` = num(r_ca)) |>
    as.data.frame()) |>
  tinytable::style_tt(j = 2:6, align = "r") |>
  tinytable::style_tt(fontsize = 0.8)
```

: The published Multi100 specification of analyst C6HJR under each treatment of the respondents who mentioned nothing. {#tbl-noop-spec}

## Multiple imputation

Imputation replaces a rule with a model. Instead of deciding what the silent would have
said, it estimates a distribution for each missing score from the respondent's
observed characteristics and draws from it, and it does so several times over so that
the uncertainty of the guess is carried into the final standard error rather than
hidden in a single filled-in number. @Dong2013Principled describe the three steps: fill
the gaps m times to produce m complete datasets, analyse each with the model you would
have run anyway, and combine the m results with the rules @Rubin1987Imputation gave,
under which the pooled estimate is the mean of the m estimates and its variance is the
mean within-imputation variance plus the between-imputation variance inflated by a
factor of 1 + 1/m. The between-imputation part is the price of not knowing, and a
single imputation, which sets it to zero, understates every standard error it touches.
How many imputations are enough depends on how much information is missing;
@Graham2007Many show that the once-standard three to five are too few when the fraction
of missing information is not small, and the course uses ten. The imputation model
should carry every variable in the analysis model and anything else that predicts
either the missing values or their missingness [@CollinsEtAl2001ComparisonInclusiveRestrictive;
@Schafer2002View], and it should not be *less* general than the analysis model, a
condition @Meng1994Imputation named congeniality and which is easy to violate here, since
the imputation happens at respondent level and the analysis at cell level.

Two engines are run on the case, because they come from different traditions and the
course should show both. **mice** [@Buuren2011Mice] imputes one variable at a time
conditional on the others, cycling until the imputations settle, and the version used
here draws each missing share from respondents with similar predicted values, which
keeps every imputed score inside the range a bounded scale can take
[@White2010Imputation]. **Amelia** [@HonakerEtAl2011AmeliaIIProgram; @KingEtAl2001AnalyzingIncompletePolitical]
fits one multivariate normal model to all the variables at once by expectation
maximisation with bootstrapping, the approach Gary King's group developed for
time-series cross-section data of exactly this shape [@HonakerKing2010WhatMissingValues],
and draws from that model; its draws can fall outside the unit interval and are clipped.
Both condition on age, sex, education and the country and year of interview. Each of the
ten completed respondent files is aggregated to country-year means and every one of the
1,560 estimand cells of the menu is fitted on each, then pooled.

```{r}
#| label: tbl-mi
#| echo: false
tinytable::tt(
  mi_spec |>
    transmute(Engine = c(mice_pmm = "mice, predictive mean matching",
                         amelia_emb = "Amelia, EM with bootstrapping")[engine],
              b = gsub("-", "−", formatC(estimate, digits = 3, format = "g"), fixed = TRUE),
              t = num(statistic, 2), df = formatC(df, format = "f", digits = 0),
              `Fraction of missing information` = num(fmi, 3)) |>
    as.data.frame()) |>
  tinytable::style_tt(j = 2:5, align = "r") |>
  tinytable::style_tt(fontsize = 0.8)
```

: The published specification under multiple imputation, pooled over ten imputations by Rubin's rules. The fraction of missing information is the share of the coefficient's uncertainty that is due to the silent respondents. {#tbl-mi}

The two engines land close to one another and close to the drop rule, which is what
should happen when the imputation model conditions on the same things that distinguish
the silent from the rest: under missing at random, filling the gaps from people who look
like them is a more careful version of leaving them out. The gap between the imputed
rows and the zero row is the part of the published result that the zero convention
supplied. That is a result about this case; @Lall2016Imputation reanalysed thirty
studies in comparative and international political economy with multiple imputation in
place of listwise deletion and found that the main results of fourteen of them
disappeared, because in that literature the missing cases were the poorer and less
democratic countries and the analyses had been run, by default, on the rest. Where the
missing are different from the observed on the thing being studied, the method of
handling them is part of the finding.

## "Don't know" as an answer

Survey methodologists have argued for fifty years about whether a "don't know" is
missing data or a response. @Francis1975Know reviewed the early evidence;
@Krosnick2001Impact showed experimentally that offering an explicit no-opinion option
does not, in general, improve the quality of the answers that remain, because many who
take it could have answered; and @Luskin2011Know found the reverse for factual
knowledge, where "don't know" mostly does mean not knowing. @Kuha2018Effect distinguish
three ways of asking, encouraging a "don't know", accepting one neutrally, and probing
it with a follow-up, and show on European Social Survey data that probing halves the
rate but changes the measurement properties of the answers it extracts, so that the
survey has traded an error of omission for an error of commission; the ESS itself does
not probe. Who says "don't know" is not random either: @Laurison2015WillingnessStateOpinion
finds the willingness to state an opinion patterned by class and education, which makes
non-response on attitude items a form of inequality in whose views reach the data.

For the case that literature cuts two ways. It supports treating a "don't know" about
the EU as non-response rather than as zero framing, which is the split rule and the
imputation option. It also warns that a group defined by not answering is not exchangeable
with the group that did, so the missing-at-random assumption behind imputation is exactly
where a critic would press. The battery's own "other" code is the small piece of good
news: `r big(n_oth)` people volunteered a meaning the list did not offer, and those
answers are opinions, not gaps.

## Sensitivity to the assumption

Nothing in the observed data can distinguish missing at random from missing not at
random; @Molenberghs2008Missingness prove that for any model of the latter there is a
model of the former that fits the observed data equally well. The response is not to
choose one but to show what the conclusion depends on. The device here is a sensitivity
band. Suppose every silent respondent would have scored what the answering respondents
in their cell scored, plus some amount δ that is the same for all of them. Then each
cell mean is the answered mean plus δ times the share of the cell that was silent, and
the published specification can be refitted for any δ. Two values are already familiar:
δ = 0 is the drop rule, and δ equal to minus the answered mean, about
`r num(delta_zero, 2)` here, is the zero rule.

```{r}
#| label: fig-delta
#| echo: false
#| fig-width: 7
#| fig-height: 3.8
#| fig-cap: "The unemployment coefficient on cosmopolitan framing, two-way fixed effects, when every silent respondent is assigned the answered mean of their cell plus δ. Zero on the horizontal axis is the drop rule; the dashed line marks the δ that reproduces the zero rule."
ggplot(band, aes(delta, t)) +
  geom_hline(yintercept = c(-1.96, 1.96), colour = "#a10000", linewidth = 0.4, linetype = 3) +
  geom_vline(xintercept = 0, colour = "#848c94", linewidth = 0.4) +
  geom_vline(xintercept = delta_zero, colour = "#0e3b6a", linewidth = 0.5, linetype = 2) +
  geom_line(colour = "#0e3b6a", linewidth = 0.9) +
  annotate("text", x = 0.02, y = max(band$t), hjust = 0, size = 3.1, colour = "#848c94",
           label = "δ = 0: drop rule") +
  annotate("text", x = delta_zero + 0.02, y = min(band$t), hjust = 0, size = 3.1,
           colour = "#0e3b6a", label = "zero rule") +
  scale_y_continuous(labels = function(x) gsub("-", "−", x, fixed = TRUE)) +
  labs(x = "δ: assumed shortfall of the silent respondents, in scale units", y = "t") +
  theme_minimal(base_size = 11)
```

The band makes the zero convention legible as what it is: the assumption that everyone
who stayed silent would have scored the lowest value the scale allows, which is the
most extreme missing-not-at-random assumption available, and the one that moves the
result furthest in the claim's direction. Whether any conclusion changes across the
band is a fact about the case, read off the figure, and it is the fact that should
accompany a result on these data. Reporting the missing-data treatment, the assumption
it makes and the sensitivity of the result to that assumption is what
@Lee2021Framework and @Sterne2009Imputation ask for in observational research, and the
proportion of missing data is not, on its own, the thing to report
[@MadleyDowd2019Proportion].

## What to take away

- A respondent who mentioned nothing is one of three people, and the survey says which.
  Look for the codes before choosing a rule.
- The original paper and its reanalyses differ in this rule and neither states it; the
  recorded coefficient rests on the convention that the silent have zero framing.
- Zero, drop and split are not robustness variants of one estimate. They change whose
  framing the average describes.
- Imputation replaces a rule with a model and carries the uncertainty into the result;
  it assumes missingness at random given what it conditions on, and that assumption is
  where a critic will press.
- Show the band. A result that survives the range of defensible assumptions about the
  silent is a different kind of result from one that depends on the zero convention.

## Exercise

1. From `teney_panel_weights.csv`, plot the no-opinion rate against the unemployment
   rate across the 270 cells and describe the pattern in one sentence. Which mechanism
   does it rule out?
2. Look up the shard for your preregistered estimand cell in `data/grid/` and compare
   its five no-opinion rows. Which rule is the zero rule closest to, and which furthest
   from?
3. Rewrite the estimand statement for the zero rule and for the drop rule. Which word
   changes?
