---
title: "Simulation as a method"
subtitle: "Fake data, known truths, and what they teach about real analyses"
bibliography: [../references.bib, ../references-new.bib]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(ggplot2)
library(lmerTest)
set.seed(20260727)

sim    <- read_csv("../data/EUframes_person_sim.csv", show_col_types = FALSE)
panel  <- read_csv("../data/EUframes_cy.csv", show_col_types = FALSE)
cal    <- read_csv("_model-outputs/sim_calibration.csv", show_col_types = FALSE)
pcoef  <- read_csv("_model-outputs/sim_calibration_person.csv",
                   show_col_types = FALSE)
truths <- read_csv("_model-outputs/sim_truths.csv", show_col_types = FALSE)
sweep  <- read_csv("_model-outputs/sim_recovery_sweep.csv", show_col_types = FALSE)
cget   <- function(nm) cal$value[cal$parameter == nm]

DIMS <- c("cosmo", "util", "comm", "lib")

# The one authored individual-level effect, transcribed from the generator; the
# demographic effects are calibrated and read from the file above.
B_UNEMPLOYED <- -0.04

# The population age the recipe centres on, and which the weight restores.
AGE_MEAN <- cget("age_mean"); AGE_SD <- cget("age_sd")
AGE_MIN  <- cget("age_min");  AGE_MAX <- cget("age_max")
AGE_REF <- AGE_MEAN + AGE_SD *
  (dnorm((AGE_MIN - AGE_MEAN) / AGE_SD) - dnorm((AGE_MAX - AGE_MEAN) / AGE_SD)) /
  (pnorm(AGE_MAX, AGE_MEAN, AGE_SD) - pnorm(AGE_MIN, AGE_MEAN, AGE_SD))

sim <- sim |>
  mutate(age10 = (age - 48) / 10,
         edu4 = factor(edu4, levels = c("le15", "e16_19", "e20plus", "studying")))

# The specification the twin is built to reproduce - the day's own anchor.
anchor <- \(y) as.formula(
  paste0(y, " ~ unemp + growth + factor(year) + factor(cntry)"))

# Dispersion measures used by the null-world and design sections, derived here
# rather than pasted so that regenerating the twin cannot leave them stale.
m_cells   <- lmer(mcosmo ~ unemp + (1 | cntry), data = panel)
SD_CNTRY  <- as.data.frame(VarCorr(m_cells))$sdcor[1]
SD_CY     <- sigma(m_cells)
SD_PERSON <- sim |>
  summarise(v = var(cosmo), .by = c(cntry, year)) |>
  summarise(sd = sqrt(mean(v))) |> pull(sd)
MEAN_COSMO <- mean(sim$cosmo)
B_COSMO <- truths$b_unemp[truths$outcome == "cosmo"]
```

This is the simulation module of the [companion curriculum](index.qmd), and it is optional. Every other quantitative page on this site analyses data that came from the world. This module works the other way round: it analyses data that came from a recipe we wrote ourselves, so that for once the true answer is known before any model runs. That inversion – fake-data simulation – is one of the core habits of modern applied statistics. *Regression and Other Stories* uses it constantly, on the argument that fitting models to data with known truths is how you learn what your models can and cannot recover [@GelmanEtAl2020RegressionOtherStories]; the same habit anchors classroom practice [@GelmanVehtari2024ActiveStatisticsStories] and the workflow literature, where experimentation on simulated data is a named pillar of serious model-building [@GelmanEtAl2026BayesianWorkflow].

The module teaches from a dataset built for the purpose: `EUframes_person_sim.csv`, a simulated person-level companion to the EU-frames case – `r format(nrow(sim), big.mark = ",")` synthetic respondents across the real 270 country-year cells, with the real macro variables attached. It is built to carry the original analysis rather than a cut-down version of it: all four framing dimensions and both composites, each with its own response to unemployment and to growth, plus the individual controls the original author used. Its [codebook](../data/EUframes_person_sim_codebook.md) states the construction in full, and two facts about it govern everything below.

First, **no Eurobarometer respondent appears in it**. The raw microdata behind the case are GESIS-licensed and cannot be redistributed in any form, including as samples or extracts – [the repositories module](repositories.qmd#sec-gesis) explains that regime – so a person-level dataset this site can legitimately ship has to be synthetic. What the licensed file supplied, read once and locally, is a set of summaries: the demographic marginals, how a respondent's own age, sex and education move their framing, and how much individuals differ within a country-year. Those are aggregate statistics of the kind published appendices carry, and they are committed as `_model-outputs/sim_calibration*.csv`.

Second – and this is the pedagogical point rather than a limitation – **the recipe is known**. A data-generating process is the explicit rule by which every value was produced. Real data never comes with its rule attached; simulated data is nothing but the rule, which is what makes it a laboratory.

## 1. The recipe, stated in full {#sec-dgp}

The generator is `_model-outputs/simulate_person_twin.R`, a committed script worth reading – it is the data-generating process made executable. Its central move is to simulate the *instrument* rather than the four scores. In the real survey a respondent hears a list of thirteen things the EU might mean and names those that apply; their dimension scores are the shares of their mentions falling to each dimension. The simulation does the same: decide whether the respondent mentions anything, then divide their mentions.

The expected composition for person $i$ in country $c$ and year $t$ has two layers. The cell layer comes from the public panel, fitted under the specification the workshop's own anchor uses:

$$
a_j + b_j\,\text{unemp}_{ct} + c_j\,\text{growth}_{ct} + \text{year}_{jt} + u_{jc} + u_{jct},
$$

and to it are added the shifts a respondent's own age, sex, education and employment status produce, calibrated from the real respondents and centred on the population marginals.

The property that matters most is what the subscript $j$ is doing there. **Every dimension has its own coefficients.** Nothing is derived from a privileged dimension with the others absorbing the remainder – which would impose the false claim that utilitarian, communitarian and libertarian framings have no independent relationship with the economy. The four shares still sum to one, but that comes free: because they sum to one in the data, a set of coefficients estimated the same way on all four sums to zero automatically. The constraint is satisfied by estimating honestly, not by holding anything passive.

Two further construction details are worth knowing. Mentioning anything at all is modelled on the logit scale, because at a base rate near six per cent a linear education effect would drive the probability below zero for most graduates – the classic linear-probability failure, met here in a setting where you can watch it. And the country and cell deviations are drawn fresh, then residualised against the full design including year, because growth is dominated by common shocks: 2009 collapsed everywhere at once, so a draw made orthogonal to raw growth can still correlate with it once year is absorbed.

```{r}
#| label: glimpse
sim |>
  slice_sample(n = 6) |>
  select(cntry, year, w1, age, edu4, unemployed, cosmo, util, comm, lib, unemp) |>
  arrange(cntry, year)
```

```{r}
#| label: identities
# The share columns are stored rounded to six decimals, so four of them can
# drift from their sum: the identities need a tolerance, not ==.
identities <- sim |>
  filter(cosmo + util + comm + lib > 0) |>
  summarise(
    `dimensions sum to 1` = all(abs(cosmo + util + comm + lib - 1) < 5e-6),
    `pos + neg = 1`       = all(abs(pos + neg - 1) < 5e-6),
    `share exactly 0`     = mean(cosmo == 0),
    `share exactly 1`     = mean(cosmo == 1))
identities
```

Among respondents who mention anything, `r sprintf("%.0f%%", 100 * identities[["share exactly 0"]])` score exactly zero on cosmopolitan framing and `r sprintf("%.0f%%", 100 * identities[["share exactly 1"]])` score exactly one – the latter because someone who mentions two items and finds both cosmopolitan scores 1 by construction. This is faithful to the real instrument, and it is the concrete reason beta regression, which needs the outcome strictly inside the unit interval, cannot be used at person level while it can be used on country-year means. The [statistical methods module](stats-methods.qmd#sec-glm) states that constraint; here you can verify it.

## 2. Simulate, fit, recover {#sec-recover}

The first thing to do with data whose truth you know is to ask a model to find it. Each dimension's true coefficient is the one the **real panel** yields under the same specification, so a model fitted to the twin should land where the same model fitted to the real data lands.

```{r}
#| label: recover
sim_cells <- sim |>
  summarise(across(c(all_of(DIMS), pos, neg), mean),
            unemp = first(unemp), growth = first(growth),
            .by = c(cntry, year))

map(truths$outcome, \(v) {
  cf <- coef(lm(anchor(v), data = sim_cells))
  tibble(outcome = v, sim_unemp = cf[["unemp"]], sim_growth = cf[["growth"]])
}) |>
  list_rbind() |>
  left_join(truths, by = "outcome") |>
  select(outcome, truth_unemp = b_unemp, sim_unemp,
         truth_growth = b_growth, sim_growth) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Read the substance first, before the arithmetic. Rising unemployment pushes framing away from **both** cosmopolitan and utilitarian terms and towards communitarian ones, while libertarian framing barely moves; growth works differently again, lifting utilitarian framing most of all – which makes sense, since the utilitarian dimension is built from the economic-prosperity and social-protection items. That pattern is not an artefact of the simulation. It is what the real panel says, carried into the twin so that a participant analysing simulated respondents meets the same structure a participant analysing the real ones would.

Now the arithmetic. The simulated coefficients sit near the truths but not on them, and the reason is worth dwelling on: each of the twin's cell means is an average over a tenth of the real respondents, so it is noisier than the panel's own. A cell-level slope here carries a standard error near a quarter of the anchor's size. That is a property of the design, not a defect, and it is the sort of thing single-dataset analysis can never show you. Repeated draws can.

```{r}
#| label: fig-sweep
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "Twenty independent draws from the same recipe, each refitted with the anchor specification. Red lines mark the truths - the real panel's own coefficients."
ggplot(sweep, aes(sim_unemp, outcome)) +
  geom_point(alpha = 0.45, colour = "#00205b", size = 1.6) +
  geom_point(data = truths, aes(b_unemp, outcome), colour = "#c8102e",
             shape = 124, size = 8) +
  labs(x = "estimated unemployment slope", y = NULL) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: sweep-summary
sweep_summary <- sweep |>
  summarise(draws = n(), mean_estimate = mean(sim_unemp),
            sd_estimate = sd(sim_unemp), .by = outcome) |>
  left_join(truths |> select(outcome, truth = b_unemp), by = "outcome") |>
  mutate(bias = mean_estimate - truth,
         bias_in_ses = bias / (sd_estimate / sqrt(draws)))
sweep_summary |> mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Across draws the estimates centre close to the truths. The last column is the test: it expresses each remaining gap as a multiple of the standard error of the twenty-draw mean. Five of the six come in under `r sprintf("%.1f", sort(abs(sweep_summary$bias_in_ses), decreasing = TRUE)[2])`, which is the size sampling alone produces; the cosmopolitan dimension sits at `r sprintf("%.1f", max(abs(sweep_summary$bias_in_ses)))`, at the edge of what twenty draws explain, so a small residual from the simplex projection cannot be ruled out there and the recipe should be read as very nearly rather than exactly unbiased. Stating that is the point of running the sweep at all: a claim of unbiasedness is only worth what the measurement behind it is worth. And any *one* draw can sit a good distance off. The committed file is one such draw, and it was kept rather than reseeded: rerolling until the estimate looked good would be the simulation-study version of the very practice this workshop is about, and that temptation has a literature of its own [@PawelEtAl2024PitfallsPotentialsSimulation].

::: {.callout-tip}
## Try it

Open the generator and change `SIM_UNEMPLOYED_EFFECT`, the one authored effect in the file, to zero. Rerun and refit. Then look at what moved and what did not: the cell-level slopes barely shift, because that effect reaches them only through the small share of respondents who are unemployed. You have just run a designed simulation study – one parameter moved, everything else held – and measured a compositional channel.
:::

## 3. Six outcomes, one instrument {#sec-outcomes}

One of the workshop's sharpest empirical findings is that on the grid as estimated, the choice of outcome – which framing scale you treat as 'EU framing' – moves results more than any other analytical decision. The twin lets you see part of why, because the dimensions are shares of one instrument and cannot move independently.

```{r}
#| label: outcome-spread
truths |>
  mutate(ratio_to_cosmo = b_unemp / b_unemp[outcome == "cosmo"],
         across(where(is.numeric), \(x) signif(x, 3)))
```

Two things sit in that table. The signs split, but **not** along the positive/negative line the day's claim-alignment convention describes: `util` is a positive framing, and half of `pos`, yet its truth shares a sign with `cosmo` on unemployment and with the negative framings on growth. The split runs between whichever dimension is gaining and whichever are giving way, and which is which depends on the predictor. What claim alignment actually rests on is narrower and does hold exactly here: `pos` and `neg` partition the mentions, so `neg` = 1 − `pos` for every mentioner and reverse-coding the negative composite is exact arithmetic.

And the magnitudes span a wide range, so an analyst choosing `util` and one choosing `comm` are estimating genuinely different quantities from one instrument. That is real, not a simulation artefact – it is the panel's own structure – and it is a substantial part of why the outcome fork carries so much specification variance. [The multiverse module](multiverse.qmd#sec-auspurg) takes that argument further, because it is precisely where Auspurg's measurement-nonequivalence objection bites.

## 4. Composition or context? {#sec-composition}

A country-year's unemployment rate could move EU framing in two different ways. It might be that unemployed *people* frame the EU differently, so a high-unemployment cell contains more of them and its mean shifts – a compositional effect. Or it might be that living through high unemployment changes how everyone frames the EU – a contextual effect. The two are indistinguishable in the country-year panel, because it contains no individuals. With person-level data they separate, and with simulated data you can check the separation against a known answer.

```{r}
#| label: composition
m_ctx <- lmer(cosmo ~ unemp + unemployed + (1 | cntry) + (1 | cntry:year),
              data = sim)
parameters::model_parameters(m_ctx, ci_method = "satterthwaite",
                             effects = "fixed")
```

The individual effect is `r sprintf("%.2f", B_UNEMPLOYED)` – around ten times the cell-level coefficient per unit – and yet it explains almost none of the cell-level relationship, because a one-point rise in a cell's unemployment rate adds only about 0.6 percentage points of unemployed respondents. Context dominates composition here because the recipe was written that way, but the arithmetic that makes a large individual effect contribute so little is general, and it is the arithmetic a reader needs whenever a paper explains an aggregate association by "it must be the people".

## 5. Recovering the individual effects {#sec-individual}

The demographic variables are not decoration either: age, sex and education carry the effects they carry in the real respondents, which is what lets the twin support the original author's own model rather than a simplified one. Education is the largest effect in the file.

```{r}
#| label: individual
m_ind <- lm(cosmo ~ age10 + female + edu4 + unemployed + factor(cntry) +
              factor(year), data = filter(sim, cosmo + util + comm + lib > 0))
ind <- parameters::model_parameters(m_ind, keep = "^(age10|female|edu4|unemp)")

pcoef |>
  filter(outcome == "cosmo") |>
  select(term, calibrated = value) |>
  add_row(term = "unemployed", calibrated = B_UNEMPLOYED) |>
  left_join(ind |> as_tibble() |> select(term = Parameter, recovered = Coefficient,
                                         se = SE),
            by = "term") |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Graduates sit about 0.12 higher on cosmopolitan framing than those who left school by 15, and the twin returns that. Notice how much more precisely these individual-level coefficients are recovered than the macro ones two sections ago: there are forty thousand respondents carrying information about education and only 270 cells carrying information about unemployment. The same dataset is richly informative about one question and thin on the other, which is the design lesson section 7 makes explicit.

::: {.callout-tip}
## Try it

Build your own placebo. Take any variable in the file, shuffle it within country-year, and refit the model above with the shuffled version in place of the real one. Its coefficient should land near zero with an interval covering it – and if you repeat that twenty times, about one run in twenty will not, which is the five-per-cent rule doing exactly what it advertises. Meeting a false positive somewhere you are certain there is nothing to find is worth more than reading about one.
:::

## 6. The degrees-of-freedom story, run live {#sec-df}

The statistical methods module tells a story it can only display, because it rests on the licensed file: a naive analysis that ignores clustering reports an enormous *t* on hundreds of thousands of degrees of freedom, while the multilevel model reports a modest one on a couple of hundred – because the predictor varies only at the country-year level. On the twin you can run the contrast.

```{r}
#| label: df-story
m3     <- lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year), data = sim)
naive  <- lm(cosmo ~ unemp, data = sim)
m3_par <- parameters::model_parameters(m3, ci_method = "satterthwaite")

tibble(
  model = c("naive OLS, clustering ignored", "three-level random intercepts"),
  estimate = c(coef(naive)[["unemp"]], fixef(m3)[["unemp"]]),
  se = c(sqrt(diag(vcov(naive)))[["unemp"]],
         m3_par$SE[m3_par$Parameter == "unemp"]),
  t = estimate / se,
  df = c(df.residual(naive), m3_par$df_error[m3_par$Parameter == "unemp"])
) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

The naive row claims forty thousand independent witnesses; the multilevel row counts the few hundred effective ones, because respondents inside a country-year share one unemployment value and largely echo their cell. Person-level sample sizes advertise precision the design does not possess whenever the predictor lives at a higher level, and the model that respects the structure pays for that respect in degrees of freedom rather than in the estimate. This is also why, in Block 1 of the workshop, two Multi100 analysts fitting person-level models to the same claim could reach such different strengths of evidence.

## 7. Weights that matter {#sec-weights}

Survey weights are usually taught as a mechanic. The interesting question is when they change anything, and simulated data can answer it because the distortion the weight corrects can be built in deliberately. The sample here was drawn with a fieldwork distortion – younger respondents over-selected – and `w1` is the exact reciprocal of that selection factor, normalised to mean 1 within each country-year.

```{r}
#| label: weights
tibble(
  quantity = c("mean age", "mean cosmopolitan share"),
  unweighted = c(mean(sim$age), mean(sim$cosmo)),
  weighted = c(weighted.mean(sim$age, sim$w1),
               weighted.mean(sim$cosmo, sim$w1)),
  population = c(AGE_REF, mean(panel$mcosmo))
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

The achieved sample is about four years too young, and weighting recovers the population age almost exactly. Because age genuinely affects framing in the recipe, that composition error propagates into the outcome, and the weight removes it. That is the whole of what a weight does to a mean: it re-balances who is counted, and it matters exactly to the extent that the mis-counted people differ on the thing being measured.

The distortion here is constant across cells, so it shifts levels without touching the relationship of interest – the common case, and the reason analysts often find weighting changes their descriptive statistics more than their coefficients. It stops being the common case the moment the distortion varies with the predictor.

::: {.callout-tip}
## Try it

In the generator, set `SIM_DISTORT_SLOPE` to 0.01 and rerun. The youth over-selection now grows with a cell's unemployment rate, from about three years' worth of age shift in the calmest labour markets to eleven in the most stressed. Refit the anchor model weighted and unweighted, and watch the unweighted slope drift off the truth while the weighted one stays on it. (Much above 0.03 the tilt turns negative in low-unemployment countries, over-selecting the old there instead, which muddies the lesson.)
:::

## 8. Building the nothing {#sec-null}

The [inference reading](../exercise/reading-inference.qmd) argues that to test a specification curve you must know what the curve would look like if there were nothing to find, and the test designed there builds that nothing by permutation. Simulation offers a more direct route: write the recipe with the effect removed.

```{r}
#| label: null-world
# One null world: the twin's cell structure with the slope written out of the
# recipe, and a small menu of estimators fitted to it. Every estimate below is
# of a quantity that is exactly zero.
simulate_null <- function(seed) {
  set.seed(seed)
  cells <- panel |>
    select(cntry, year, n_cy, unemp) |>
    mutate(n_sim = pmax(1L, as.integer(round(n_cy * 0.10))),
           u_cy = rnorm(n(), 0, SD_CY))
  cells |>
    left_join(cells |> distinct(cntry) |>
                mutate(u_c = rnorm(n(), 0, SD_CNTRY)), by = "cntry") |>
    uncount(n_sim) |>
    mutate(cosmo = pmin(pmax(MEAN_COSMO + 0 * unemp + u_c + u_cy +
                               rnorm(n(), 0, SD_PERSON), 0), 1))
}

d_null <- simulate_null(2026)
list(
  pooled      = \(d) lm(cosmo ~ unemp, data = d),
  country_re  = \(d) lmer(cosmo ~ unemp + (1 | cntry), data = d),
  three_level = \(d) lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year), data = d),
  cell_means  = \(d) lm(mcosmo ~ unemp,
                        data = summarise(d, mcosmo = mean(cosmo),
                                         unemp = first(unemp),
                                         .by = c(cntry, year)))
) |>
  imap(\(f, nm) {
    p <- parameters::model_parameters(f(d_null))
    tibble(estimator = nm, estimate = p$Coefficient[p$Parameter == "unemp"],
           p_value = p$p[p$Parameter == "unemp"])
  }) |>
  list_rbind() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Every row estimates a quantity known to be zero. Rerun with a different seed – the function takes one for exactly that reason – and the estimates wander, occasionally straying to nominal significance, most readily in the pooled row, which treats forty thousand echoing respondents as independent. A miniature multiverse fitted to a null world is the cleanest available intuition for what a full-scale permutation test constructs: the distribution of whole curves under nothing. It also disciplines how the real curve's percentages should be read, a point [the multiverse module](multiverse.qmd#sec-reading) develops – some share of any universe will look supportive under a true zero, so a share alone, with no null to stand it against, is a description rather than a verdict.

```{r}
#| label: fig-lineup
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "A lineup: nine cell-level scatters of mean simulated cosmopolitan framing against unemployment. Eight permute the unemployment series within country; one is the twin's real structure. Pick the panel that looks like an effect before reading on."
real_panel <- sample(1:9, 1)
map(1:9, \(i) {
  d <- if (i == real_panel) sim_cells else
    sim_cells |> mutate(unemp = sample(unemp), .by = cntry)
  d |> mutate(panel = i)
}) |>
  list_rbind() |>
  ggplot(aes(unemp, cosmo)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "#00205b") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6, colour = "#c8102e") +
  facet_wrap(~panel) +
  labs(x = "unemployment (%)", y = "cell mean, simulated cosmopolitan share") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: lineup-answer
#| echo: false
#| output: asis
cat(sprintf("::: {.callout-note collapse=\"true\"}\n## Which panel is real?\nPanel %d. If you picked it, you have informally rejected the null by eye – the logic of a permutation test, run on your visual system. If you could not, that is itself evidence about how strong a level-2 relationship of this size looks at this sample size.\n:::\n", real_panel))
```

## 9. Design by simulation {#sec-design}

Power analysis for multilevel designs rarely has clean formulas, and the standard advice is to simulate: generate data from the design you are contemplating and see how precisely it estimates the quantity you care about [@GelmanEtAl2020RegressionOtherStories; @DeBruineBarr2021UnderstandingMixedeffectsModels]. Precision on a level-2 slope is bought mainly with countries and years rather than with respondents, so it is worth asking what four times as many of each would be worth.

```{r}
#| label: design
simulate_design <- function(k, seed, per_cell = 1) {
  set.seed(seed)
  cells <- panel |>
    select(cntry, year, n_cy, unemp) |>
    crossing(clone = seq_len(k)) |>
    mutate(cntry = paste0(cntry, "_", clone),
           n_sim = pmax(1L, as.integer(round(n_cy * 0.10 * per_cell))),
           u_cy = rnorm(n(), 0, SD_CY))
  cells |>
    left_join(cells |> distinct(cntry) |>
                mutate(u_c = rnorm(n(), 0, SD_CNTRY)), by = "cntry") |>
    uncount(n_sim) |>
    mutate(cosmo = pmin(pmax(MEAN_COSMO + B_COSMO * unemp + u_c + u_cy +
                               rnorm(n(), 0, SD_PERSON), 0), 1))
}

list(c(k = 1, per_cell = 1), c(k = 1, per_cell = 4), c(k = 4, per_cell = 1)) |>
  map(\(d) {
    m <- lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year),
              data = simulate_design(d[["k"]], seed = 20260727 + d[["k"]],
                                     per_cell = d[["per_cell"]]))
    p <- parameters::model_parameters(m, ci_method = "satterthwaite")
    tibble(countries = 27 * d[["k"]], respondents = nobs(m),
           se = p$SE[p$Parameter == "unemp"],
           df = p$df_error[p$Parameter == "unemp"])
  }) |>
  list_rbind() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Read the three rows against each other. Quadrupling the respondents inside the existing map does buy precision – people are not worthless – but it buys much less than quadrupling the countries, and it buys almost no degrees of freedom, because the number of units carrying independent information about unemployment has not changed. No feasible amount of extra survey data inside the existing 27 countries would do what 81 more countries would, and no choice on the specification menu can manufacture level-2 information the design does not contain. That is the deep reason this literature argues within wide uncertainty bands, and simulation measures a design's ceiling before anyone commits to it. A methodological literature exists on doing this properly – declaring aims, data-generating mechanisms, estimands, methods and performance measures rather than tinkering [@MorrisEtAl2019UsingSimulationStudies] – and it is the natural next reading.

## Where this leads {#sec-further}

Four routes onward. For the habit itself, *Regression and Other Stories* is the anchor text, its fake-data simulations short, constant and cumulative [@GelmanEtAl2020RegressionOtherStories], with *Active Statistics* packaging the same moves as classroom activities [@GelmanVehtari2024ActiveStatisticsStories]. For multilevel structures specifically, DeBruine and Barr's tutorial is sections 2 and 6 of this module done properly: it builds a mixed-effects simulation from nothing and recovers its parameters, on the argument that constructing the data is how you come to understand the model [@DeBruineBarr2021UnderstandingMixedeffectsModels]. For simulation inside a modelling workflow – prior and posterior predictive checking, calibration, experimentation on synthetic data before touching the real thing – the Bayesian workflow book is the current statement [@GelmanEtAl2026BayesianWorkflow], and its habits transfer well beyond Bayesian settings. And for simulation as a research method in its own right, the design framework of @MorrisEtAl2019UsingSimulationStudies supplies the vocabulary, the book-length treatment written for social scientists is @CarseyHarden2014MonteCarloSimulation, and an applied Monte Carlo study this site already cites, on cross-lagged panel models, shows the genre working on a question adjacent to this case's own estimator fork [@LeszczenskyWolbring2022HowDealReverse].

One closing thought ties the module back to the rest of the site. Simulation studies are analyses too, and they have researcher degrees of freedom of their own – conditions, performance measures, which results get reported – so a comparative study can be tuned until any method looks superior [@PawelEtAl2024PitfallsPotentialsSimulation]. The disciplines this course keeps returning to, declared designs and justified choices, apply in the laboratory exactly as they apply in the field. The generator, the calibration script, the codebook and this page's chunks are all committed, so the fastest route of all is the one from the first Try-it: change the truth, rerun, and watch what your tools do when you already know the answer.
