---
title: "Simulation as a method"
subtitle: "Fake data, known truths, and what they teach about real analyses"
bibliography: [../references.bib, ../references-new.bib]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(ggplot2)
library(lmerTest)
set.seed(20260727)
sim <- read_csv("../data/EUframes_person_sim.csv", show_col_types = FALSE)
panel <- read_csv("../data/EUframes_cy.csv", show_col_types = FALSE)
# The generator's named truth – transcribed from
# companion/_model-outputs/simulate_person_twin.R, which is the single source.
SIM_B0       <- 0.5526
SIM_B_UNEMP  <- -0.005237
SIM_SD_CNTRY <- 0.06131
SIM_SD_CY    <- 0.03853
SIM_SD_E     <- 0.36806
```

This is the simulation module of the [companion curriculum](index.qmd), and it is optional. Every other quantitative page on this site analyses data that came from the world – 270 country-years distilled from sixteen survey waves. This module works the other way round: it analyses data that came from a recipe we wrote ourselves, so that for once the true answer is known before any model runs. That inversion, variously called fake-data simulation or simulation-based learning, is one of the core habits of modern applied statistics – *Regression and Other Stories* uses it constantly, on the argument that fitting models to data with known truths is how you learn what your models can and cannot recover [@GelmanEtAl2020RegressionOtherStories], and the same habit anchors classroom practice [@GelmanVehtari2024ActiveStatisticsStories] and the workflow literature, where experimentation on simulated data is a named pillar of serious model-building [@GelmanEtAl2026BayesianWorkflow].

The module teaches from a dataset built for exactly this purpose: `EUframes_person_sim.csv`, a simulated person-level twin of the EU-frames case – 20,829 synthetic respondents in the real 270 country-year cells, with the real macro variables attached. Its [codebook](../data/EUframes_person_sim_codebook.md) states the construction in full. Two facts about it govern everything below. First, **no Eurobarometer respondent appears in it**: the raw microdata behind the case are GESIS-licensed and cannot be redistributed in any form, including as samples or extracts ([the repositories module](repositories.qmd#sec-gesis) explains that regime), so a person-level dataset this site can legitimately ship has to be synthetic, generated from disclosed aggregate parameters only. Second – and this is the pedagogical point, not a limitation – **its data-generating process is known and named**. A data-generating process (DGP for short) is the explicit recipe by which every value in a dataset was produced: here, a person's simulated cosmopolitanism score is a country effect plus a country-year effect plus individual noise, hung on a linear response to the real unemployment series. Real data never comes with its recipe. Simulated data is nothing but the recipe, and that is what makes it a laboratory.

## 1. The recipe, stated in full {#sec-dgp}

The twin is generated by `companion/_model-outputs/simulate_person_twin.R`, a short committed script you can and should read – it is the DGP made executable. In model notation, for person $i$ in country $c$ and year $t$:

$$
\text{cosmo}_{ict} = \beta_0 + \beta_1\,\text{unemp}_{ct} + u_c + u_{ct} + e_{ict}
$$

with $\beta_0 = 0.5526$, $\beta_1 = -0.005237$, $u_c \sim N(0, 0.06131^2)$, $u_{ct} \sim N(0, 0.03853^2)$ and $e_{ict} \sim N(0, 0.36806^2)$, the result clipped to the outcome's [0, 1] bounds. Those five constants are not invented: they are the fixed effects and variance components of the three-level model that [the statistical methods module](stats-methods.qmd#sec-multilevel) displays fitted to the real 416,698 respondents – a single teaching fit whose aggregate parameters are public. So the twin is the real analysis run *forwards*: we took the model that summarised the world and asked it to produce a world of its own.

```{r}
#| label: glimpse
sim |>
  slice_sample(n = 6) |>
  arrange(cntry, year)
```

Three warnings before anything is fitted, all stated in the codebook and worth internalising as habits. The demographic columns (`age`, `female`, `edu4`) and the weight (`w1`) are stylised marginals, drawn independently of the outcome – any 'effect' they show is sampling noise, and section 5 turns that into an exercise rather than a trap. The macro columns (`unemp`, `growth`, `bailout`) are real published aggregates, joined from the public panel. And the clipping step means the observed outcome is a *censored* version of the latent recipe – about 18% of latent draws fall outside [0, 1] and get pinned to the boundary – which will matter in a moment.

## 2. Simulate, fit, recover {#sec-recover}

The first thing to do with data whose truth you know is ask your model to find it. This is parameter recovery: fit the model that matches the DGP and check the estimate against the named constant.

```{r}
#| label: recover
m3 <- lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year), data = sim)
summary(m3)
```

Read the unemployment row against the truth of $\beta_1 = -0.005237$. The estimate will not equal it – this is one draw of a random recipe, and with only 270 country-year cells carrying the identifying variation, the Monte-Carlo error on a level-2 coefficient is visible – but the confidence interval should cover it comfortably:

```{r}
#| label: recover-ci
confint(m3, parm = "unemp", method = "Wald") |> round(4)
```

There is a second, subtler lesson in the recovery. The point estimate sits closer to zero than the latent-scale truth by a factor of roughly 0.84, and that shrinkage is not sampling error – it is the clipping. Bounding a Gaussian outcome to [0, 1] pins the extreme draws to the boundary, compresses the residual spread (the fitted residual SD is about 0.31 against the recipe's 0.368), and attenuates a linear slope. You have met this phenomenon before in different clothing: it is the reason the specification menu carries an outcome-family axis, where a logit-link beta model replaces the linear one precisely because bounded outcomes and linear responses sit awkwardly together. The generator's own validation report quantifies the attenuation exactly, by fitting the same model to the unclipped latent outcome (which recovers the truth with no bias) and to the clipped one. Simulation is the only setting in which you can *see* attenuation this cleanly, because only here do you hold both the latent truth and the bounded observation.

::: {.callout-tip}
## Try it

Open `simulate_person_twin.R`, change `SIM_B_UNEMP` to twice its value, rerun it, and refit the recovery model on the regenerated file. Then set it to exactly zero and refit again. You have just run your first designed simulation study: one parameter moved, everything else held, and the estimator's behaviour observed at a truth of your choosing.
:::

## 3. The degrees-of-freedom story, run live {#sec-df}

The statistical methods module tells a story it can only *display*: on the real 416,698 respondents, a naive analysis that ignores clustering reports t = −24.5 on roughly 391,000 degrees of freedom, while the three-level model reports t = −6.5 on about 253 – because the unemployment series varies at the country-year level, so there are only 270 informative units, not 416,698. On the twin you can run the whole contrast yourself.

```{r}
#| label: df-story
naive  <- lm(cosmo ~ unemp, data = sim)
m3_par <- parameters::model_parameters(m3, ci_method = "satterthwaite")

tibble(
  model = c("naive OLS, clustering ignored", "three-level random intercepts"),
  estimate = c(coef(naive)["unemp"], fixef(m3)["unemp"]),
  se = c(sqrt(diag(vcov(naive)))["unemp"],
         m3_par$SE[m3_par$Parameter == "unemp"]),
  t = estimate / se,
  df = c(df.residual(naive), m3_par$df_error[m3_par$Parameter == "unemp"])
) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

The two rows report nearly the same slope – ignoring clustering does not bias this estimate – but disagree about certainty. The naive row claims twenty thousand independent witnesses; the multilevel row correctly counts the ~240 effective ones, because respondents inside a country-year share the same unemployment value and largely echo each other's cell. Both facts matter for reading published research: person-level sample sizes advertise precision that the design does not possess whenever the predictor lives at a higher level, and the model that respects the structure pays for that respect in degrees of freedom, not in the estimate. This is also why, in Block 1 of the workshop, two Multi100 analysts fitting person-level models could reach such different strengths of evidence from the same claim – the degrees-of-freedom convention *is* an analytical choice, made visible here because the truth is known.

## 4. Building the nothing, hands-on {#sec-null}

The [inference reading](../exercise/reading-inference.qmd) argues that to test a specification curve you must know what the curve would look like if there were nothing to find – and the workshop's registered test builds that 'nothing' by permutation. Simulation gives you a more direct route to the same object: write a recipe with the effect removed. Set $\beta_1 = 0$, keep every other part of the structure – the country effects, the cell effects, the person noise, the cell sizes – and generate a world in which unemployment does nothing.

```{r}
#| label: null-world
# One null draw: the twin's structure with the slope written out of the recipe.
simulate_null <- function(seed) {
  set.seed(seed)
  cells <- panel |>
    select(cntry, year, n_cy, unemp) |>
    mutate(n_sim = pmax(1L, as.integer(round(n_cy * 0.05))),
           u_cy = rnorm(n(), 0, SIM_SD_CY))
  cells |>
    left_join(cells |> distinct(cntry) |>
                mutate(u_c = rnorm(n(), 0, SIM_SD_CNTRY)),
              by = "cntry") |>
    uncount(n_sim) |>
    mutate(cosmo = pmin(pmax(
      SIM_B0 + 0 * unemp + u_c + u_cy + rnorm(n(), 0, SIM_SD_E), 0), 1))
}

# A miniature multiverse on one null world: four estimators of a truth of zero.
null_world <- simulate_null(2026)
null_fits <- list(
  pooled      = \(d) lm(cosmo ~ unemp, data = d),
  country_re  = \(d) lmer(cosmo ~ unemp + (1 | cntry), data = d),
  three_level = \(d) lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year),
                          data = d),
  cell_means  = \(d) d |>
    summarise(mcosmo = mean(cosmo), unemp = first(unemp),
              .by = c(cntry, year)) |>
    (\(a) lm(mcosmo ~ unemp, data = a))()
)
null_fits |>
  imap(\(f, nm) {
    m <- f(null_world)
    p <- parameters::model_parameters(m)
    tibble(estimator = nm,
           estimate = p$Coefficient[p$Parameter == "unemp"],
           p_value = p$p[p$Parameter == "unemp"])
  }) |>
  list_rbind() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Every row of that table is an estimate of a quantity you *know* to be zero. Run it a few times with different seeds (the chunk takes a seed argument for exactly that reason) and you will see the estimates wander, occasionally straying to nominal significance – most readily in the pooled row, which treats twenty thousand echoing respondents as independent. A mini-multiverse fitted to a null world is the cleanest possible intuition for what the registered permutation test constructs at full scale: the distribution of whole curves under nothing. It also disciplines how you read the real curve's percentages, a point [the multiverse module](multiverse.qmd#sec-reading) develops: some share of any universe will look supportive under a true zero, so a share alone, without a null to stand it against, is a description and not a verdict.

## 5. What noise looks like {#sec-noise}

The codebook promises that the demographic columns carry no information about the outcome – they were drawn independently. That makes them a built-in placebo bench: any association a model finds between them and `cosmo` is, by construction, noise.

```{r}
#| label: placebo
lmer(cosmo ~ female + age + (1 | cntry) + (1 | cntry:year), data = sim) |>
  parameters::model_parameters(ci_method = "satterthwaite") |>
  print()
```

With around twenty thousand observations and a truth of exactly zero, the estimates come out tiny and the intervals cover zero – this is what a true nothing looks like at scale, and it is worth staring at. When a person-level dataset this size hands you a 'significant' coefficient of comparable magnitude in real research, the placebo bench is the right mental reference: the machinery will happily estimate patterns this small, and only design, theory and – as in Lakens and colleagues' argument in the multiverse module – warranted assumptions separate an effect from an accident. Statistics education increasingly teaches this comparison visually: the lineup protocol embeds the plot of real data among plots of pure noise and asks whether you can pick it out [@Loy2021BringingVisualInference].

```{r}
#| label: fig-lineup
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "A lineup: nine cell-level scatters of mean simulated cosmopolitanism against unemployment. Eight permute the unemployment series within country; one is the twin's real structure. Before reading on, pick the panel that looks like an effect."
cell_means <- sim |>
  summarise(mcosmo = mean(cosmo), unemp = first(unemp), .by = c(cntry, year))
real_panel <- sample(1:9, 1)
lineup <- map(1:9, \(i) {
  d <- if (i == real_panel) cell_means else
    cell_means |> mutate(unemp = sample(unemp), .by = cntry)
  d |> mutate(panel = i)
}) |> list_rbind()
ggplot(lineup, aes(unemp, mcosmo)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "#00205b") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6, colour = "#c8102e") +
  facet_wrap(~panel) +
  labs(x = "unemployment (%)", y = "cell mean, simulated cosmo") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: lineup-answer
#| echo: false
#| output: asis
cat(sprintf("::: {.callout-note collapse=\"true\"}\n## Which panel is real?\nPanel %d. If you picked it, you have informally rejected the null by eye – the logic of a permutation test, run on your visual system. If you could not tell, that is itself evidence about how strong this level-2 relationship is at 5%% scale.\n:::\n", real_panel))
```

## 6. Design by simulation: why 27 countries is the constraint {#sec-design}

Power analysis for multilevel designs rarely has clean formulas, and the standard advice is to simulate: generate data from the design you are contemplating and see how precisely it estimates the quantity you care about [@GelmanEtAl2020RegressionOtherStories]. The twin makes one such comparison vivid. Level-2 precision here is bought with countries and years, not with respondents – so what would this design's precision look like if Europe had four times as many countries?

```{r}
#| label: design
# The same recipe on an inflated map: each country cloned k times with fresh
# country and cell effects, respondents unchanged per cell. Precision at
# level 2 should improve with the square root of the cluster count.
simulate_design <- function(k, seed) {
  set.seed(seed)
  cells <- panel |>
    select(cntry, year, n_cy, unemp) |>
    crossing(clone = seq_len(k)) |>
    mutate(cntry = paste0(cntry, "_", clone),
           n_sim = pmax(1L, as.integer(round(n_cy * 0.05))),
           u_cy = rnorm(n(), 0, SIM_SD_CY))
  cells |>
    left_join(cells |> distinct(cntry) |>
                mutate(u_c = rnorm(n(), 0, SIM_SD_CNTRY)),
              by = "cntry") |>
    uncount(n_sim) |>
    mutate(cosmo = pmin(pmax(
      SIM_B0 + SIM_B_UNEMP * unemp + u_c + u_cy +
        rnorm(n(), 0, SIM_SD_E), 0), 1))
}

c(1, 4) |>
  map(\(k) {
    m <- lmer(cosmo ~ unemp + (1 | cntry) + (1 | cntry:year),
              data = simulate_design(k, seed = 20260727 + k))
    p <- parameters::model_parameters(m, ci_method = "satterthwaite")
    tibble(countries = 27 * k, rows = nobs(m),
           se = p$SE[p$Parameter == "unemp"],
           df = p$df_error[p$Parameter == "unemp"])
  }) |>
  list_rbind() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))
```

Quadrupling the countries roughly halves the standard error – the square-root law operating on clusters, not on respondents. No feasible amount of extra survey data inside the existing 27 countries would do what 81 additional countries would, and no reanalysis choice on the specification menu can manufacture level-2 information the design does not contain. That is the deep reason the EU-frames literature argues within wide uncertainty bands, and simulation is the tool that lets you *measure* a design's ceiling before committing to it. A methodological literature exists on doing this properly – planning simulation studies with declared aims, data-generating mechanisms, estimands, methods and performance measures rather than ad hoc tinkering [@MorrisEtAl2019UsingSimulationStudies] – and it is the natural next reading once the habit takes hold.

## Where this leads {#sec-further}

Four onward routes. For the habit itself, *Regression and Other Stories* is the anchor text – its fake-data simulations are short, constant and cumulative [@GelmanEtAl2020RegressionOtherStories], and *Active Statistics* packages the same moves as classroom activities [@GelmanVehtari2024ActiveStatisticsStories]. For multilevel structures specifically, DeBruine and Barr's tutorial is this module's sections 2 and 3 done properly: it builds a mixed-effects simulation from nothing and recovers its parameters, and understanding the model *by constructing its data* is the whole method [@DeBruineBarr2021UnderstandingMixedeffectsModels]. For simulation inside a modelling workflow – prior and posterior predictive checking, calibration, experimentation on synthetic data before touching the real thing – the Bayesian workflow book is the current statement [@GelmanEtAl2026BayesianWorkflow], and its habits transfer well beyond Bayesian settings. For simulation as a research method in its own right – evaluating estimators, as this module did informally in sections 2 through 6 – the design framework of @MorrisEtAl2019UsingSimulationStudies gives the vocabulary, the book-length treatment written for social scientists is @CarseyHarden2014MonteCarloSimulation, and an applied Monte-Carlo study the site already cites, on cross-lagged panel models, shows the genre working on a question adjacent to this case's own estimator fork [@LeszczenskyWolbring2022HowDealReverse].

One closing thought ties this module back to [the multiverse module](multiverse.qmd). Simulation studies are analyses too, and they have researcher degrees of freedom of their own: a comparative simulation study can be tuned – in its conditions, its performance measures, its reported subsets – until any method looks superior [@PawelEtAl2024PitfallsPotentialsSimulation]. The disciplines this course keeps returning to, declared designs and justified choices, apply to the laboratory just as they apply to the field. The generator script, the codebook and this page's own chunks are all committed, so the fastest route of all is the one from the first Try-it: change the truth, rerun, and watch what your tools do when you know the answer.
